Legacy Data Confounds Modern Genomics Studies

### 2018-09-11 legacy.tex
updated discussion, removed 'reverse GWAS'

%While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \citep{Pleasance2010}.
%Large cohorts have recently become available to increase statistical power to resolve some of these finer scale mutational signatures.

Differences in mutational signatures across human populations have recently been reported.
In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}. 
Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
Harris and Pritchard further identified distinct mutational spectra across a range of populations \citep{Harris2017a}, which were further studied in \citep{Aikens2018}

@@

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a reverse genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
Whereas the majority of GWAS use linear regression predicting a phenotype based on genotype, we are looking for loci where the genotypes are predictable based on the data quality scores.
This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in low-mapping quality individuals.

Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.

%For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 
The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
For population genetic analyses, the removal of individuals with low quality sequence data appears preferable to filtering positions.

@@

\subsection{Overlap of significant SNPs}
The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{meta-analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.  
\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}

@@

\subsection{Imputation}
The state of the art imputation servers use a combination of many databases including many that are not freely available.
From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
In order to investigate the proportion of variants that we have identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
We found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed. 
This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed. 
%\sgcomment{All of them? How is it possible if we only submitted Japanese? Also, I thought that we had only submitted a couple of chromosomes.} 
%\luke{Submitted all of the 1kGP, queried the combined meta-analysis list of SNPs}
\sgcomment{It may be useful to impute some of the NAG data.}

@@

One of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation. \sgcomment{If they used HapMap, how did they get affected by the 1000 Genomes results?}
The other study used the 1kGP sequence data and cell lines directly.
Both papers used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds. 
They also removed rare alleles and alleles with high degrees of missingness. 
Despite using the state of the art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for biological traits.

@@

%\subsection{Twins Research Human Genetics}
%\sgcomment{Had we ever discusssed that these GWAS hits were not genome-wide significant?}
This paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below significance below $ p < 10^{-5}$.
The data used was imputed genotype data using Hapmap II as the reference.
One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality \citep{Ebejer2013}.

While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified. 
Furthermore, the approach they used to impute genotype data for GWAS is very common.
%A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}

%\subsection{PLos One}
This paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
They did PCR confirmation of EBV copy number using 7 samples from some of the populations, however all of these samples have relatively high quality data.
They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
However they reported the SNPs reading significance below $ p < 10^{-6}$. 
Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) we're identified by our regression against data quality \citep{Mandage2017}.

@@

\subsection{Technical Artifact}
\sgcomment{needs some editing}
The variants identified in this study are likely to be technical artifacts from legacy technology.\citep{Minoche2011} \luke{<-citation} 
If these sequencing platforms are no longer being used then it it likely that these artifacts are no longer being actively introduced in recent sequence data.
However, because the 1kGP data is widely used as a reference database, these variants are being imputed onto new genotype data.
We have shown that these variants have been significantly associated using traditional GWAS methods.
However this is not the only method that would be affected by these false positives. 
Polygenic risk scores, meta analyses and other methods that combine the scores of multiple variants might also include some of these variants in their analyses.
The inclusion of these variants could also affect population genetics analyses that compare and contrast populations based on allele frequencies.
luke{Multiple explanations for the signal, technical artifact, cell line artifact. The presence of these SNPs in the HapMap suggests a cell line effect, however with only 1\% of sites identified in 1kGP present in HapMap it's not a smoking gun.}

\subsection{Recommendations}
\sgcomment{needs some editing}
The most conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low quality.
In this case, we used a cut-off of an average quality of mapped bases over 30.
Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test. (See supplementary material for lists of SNPs)
%\subsection{Power limits and Quality Controls}
%If these variants are truly spurious, why were they not identified by the Hardy-Weinberg test of heterozygosity?
%how many hetd do you need to detect HWE significance?
%We were able to identify many variants with frequencies below \todo{the power threshold for HWE}.

%\subsection{Technical Source of Error}
%B tail trimming  (Minoche et al. Genome Biology 2011)
%issues with PCR?
%repetitive regions, GC bias...

@@

On a technical front, we were surprised that strong correlations between SNPs and technical covariates in the 1kGP project had not been identified before. The reverse GWAS approach is rather straightforward, and should probably be a standard in a variety of -omics studies.

@@

%but evidence points to the fact that the artifact disappeared in the later stages of the 1kGP, and does not corrupt more recent data.

%As more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
%These results bring forth many questions regarding the reliability of legacy datasets. 
%Moreover, since there are so many broad applications of imputation, it frames the question for reference data turnover.

%Our method identifies spurious mutations by correlating mutations with data quality metrics. 
%We propose including our quality control methods to identify possible false positives in sequencing data. 
%We have focused on the 1000 Genomes Project dataset as its quality metrics were freely available, however the issues of quality control are not limited to this one consortium. 
%This study only used one dataset and one quality metric, but using this same approach can be used to identify more false positives in many more datasets.

@@

%\bibliographystyle{plain}
\bibliography{Legacy.bib}

>>>>

While certain mutagens have very distinct mutational signatures, others are more subtle and require much larger datasets to identify because they tend to occur less frequently and with less genomic specificity \citep{Pleasance2010}.
Differences in mutational signatures across human populations have recently been reported.
In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}. 
Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
Harris and Pritchard further identified distinct mutational spectra across a range of populations \citep{Harris2017a}, which were further studied in \citep{Aikens2018}

@@

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
Whereas the majority of GWAS use linear regression predicting a phenotype based on genotype, we are looking for loci where the genotypes are predictable based on the data quality scores.
This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in low-mapping quality individuals.

Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques. 
The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
For population genetic analyses, the removal of individuals with low quality sequence data appears preferable to filtering positions.

@@

\subsection{Overlap of significant SNPs}
The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{combined analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.  
\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}

@@

\subsection{Imputation}
The state of the art imputation servers use a combination of many databases including many that are not freely available.
From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
In order to investigate the proportion of variants that we have identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
We found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed. 
This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed. 
%\sgcomment{It may be useful to impute some of the NAG data.}

@@

One of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
The other study used the 1kGP sequence data and cell lines directly.
Both papers used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds. 
They also removed rare alleles and alleles with high degrees of missingness. 
Despite using the state of the art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for biological traits.

@@

One paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below significance below $ p < 10^{-5}$.
The data used was imputed genotype data using Hapmap II as the reference.
One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality \citep{Ebejer2013}.
While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
Finding some of these SNPs in imputed HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
We found that roughly 1\% of the SNPs associated to quality were found in HapMap,  rs6057648 is among those found in HapMap.

Another paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
They did PCR confirmation of EBV copy number using 7 samples from some of the populations, however all of these samples have relatively high quality data.
They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
However they reported the SNPs reading significance below $ p < 10^{-6}$. 
Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) we're identified by our regression against data quality \citep{Mandage2017}.

@@

\subsection{Technical Artifact}
\luke{edited}
The variants identified in this study are likely to be technical artifacts from legacy technology.
Different sequencing technologies will have different error profiles. 
A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011}.
The quality of sequencing data decreases along the length of the read.
These regions of low quality at the ends of reads are know as B-tails.
The same report found that the GAII also had much higher rates of reads with B-tails.
However, many variant calling algorithms will discard reads with high error rates \citep{Minoche2011}.
There are indeed multiple explanations for the signal. 
The presence of these SNPs in the HapMap suggests a cell line effect, however this result is not very convincing with only 1\% of sites identified in 1kGP present in HapMap.
Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.

If these sequencing platforms are no longer being used then it it likely that these artifacts are no longer being actively introduced in recent sequence data.
However, because the 1kGP data is widely used as a reference database, these variants are being imputed onto new genotype data.
We have shown that these variants have been significantly associated using traditional GWAS methods.
However this is not the only method that would be affected by these false positives. 
Polygenic risk scores, meta analyses and other methods that combine the scores of multiple variants might also include some of these variants in their analyses.
The inclusion of these variants could also affect population genetics analyses that compare and contrast populations based on allele frequencies.

\subsection{Recommendations}
\luke{edited}
While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet the current quality standards.
A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low quality.
In this case, we used a cut-off of an average quality of mapped bases over 30.
Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test. (See supplementary material for lists of SNPs)

@@

On a technical front, we were surprised that strong correlations between SNPs and technical covariates in the 1kGP project had not been identified before. 
The GWAS approach is rather straightforward, and should probably be a standard in a variety of -omics studies.

@@

@@

\bibliography{Legacy.bib}
